Adaptive Sequential Model Selection

### 2015-07-10 paper.tex
rob

\begin{table}[ht]
\centering
\begin{tabular}{rrrrr}
  \hline\\
 Step& Predictor& Naive pvalue& Saturated pvalue& Max-t pvalue\cr
  \hline\\
1 & 3& 0.00 & 0.00 & 0.00 \\ 
  2 & 9 & 0.00 & 0.00 & 0.00 \\ 
  3 & 4 & 0.00 & \color{red}{0.05} & 0.00 \\ 
  4 & 20 & 0.00 & 0.33 & 0.01 \\ 
  5 & 37 & 0.00 & 0.76 & 0.05 \\ 
  6 & 7 & 0.00 & 0.25 & 0.05 \\ 
  7 & 2 & 0.00 & 0.00 & 0.00 \\ 
  8 & 19 & 0.02 & 0.03 & \color{red}{0.37} \\ 
  9 & 11 & 0.11 & 0.55 & 0.95 \\ 
  10 & 52 & 0.21 & 0.91 & 0.92 \\ 
  11 & 57 & 0.05 & 0.37 & 0.21 \\ 
  12 & 24 & 0.22 & 0.15 & 0.00 \\ 
  13 & 5 & 0.18 & 0.07 & 0.10 \\ 
  14 & 49 & 0.16 & 0.97 & 0.75 \\ 
  15 & 30& 0.21 & 0.15 & 0.05 \\ 
  16 & 46 & \color{red}{0.16} & 0.05 & 0.53 \\ 
  17 & 59 & 0.27 & 0.45 & 0.68 \\ 
  18 & 33 & 0.12 & 0.71 & 0.80 \\ 
  19 & 62 & 0.08 & 0.40 & 0.48 \\ 
  20 & 27 & 0.17 & 0.60 & 0.50 \\ 
   \hline\\
\end{tabular}
\caption[tab:diab]{\em Forward stepwise regression for the diabetes data: naive p-values from forward stepwise regression, 
p-values from the ``saturated'' model, and the ``max-t'' p-values that are presented in this paper. The red annotation indicates the stopping step
prescribed by {\tt forwardStop}. For the max-t p-values, this gives exact FDR control at the 10\% level.}
\label{tab:diab}
\end{table}
The table shows the naive p-values from forward stepwise regression, that do not account for the selection, as well as the
p-values from the ``saturated'' model, and the ``max-t'' p-values that are presented in this paper.
The latter two sets of p-values test someone different hypotheses:  the p-values from the saturated model test whether the partial regression of the predictor just added is zero,
what we call the ``incremental null''
In contrast, the max-t pvalues test whether the current model is adequate, that is, all predictors not in the model  have coefficients equal to zero,
the ``complete null''. (The issue of what hypothesis to test is discussed in section \ref{sec:whichnull}.)
Now it turns out that the max-t test if often more powerful that the saturated model test, and importantly, the max-t pvalues are exactly
independent after we reach the point the model sequence when the complete null is true. This independence means that we can apply
sequential FDR exact stopping rules, such as ``forwardStop'' (\citet{gsell2013sequential}). When applied to the max-t pvalues at FDR 0.10, {\tt forwardStop}
suggested stopping at step 8. In constant, the saturated lvalues gave stopping a point of 3, with no guarantee of FDR control
due to the lack of independence of the p-values.

@@

By contrast, the LASSO path does not in general add the variable with the largest $t$-statistic. In that case, we could use instead the {\em selective next-$\lambda$ test}, which uses test statistic
\begin{equation}
  U_k = \inf\{ \lambda:\; \hat\beta^\lambda \notin \beta(E_{k-1})\}.
\end{equation}
Here $\beta(E_{k-1})$ is the support of the ever-active set $E_{k-1}$, so that $U_k$ is the next $\lambda$ at which a new predictor is added by the lasso.
\RTcomment{Will: we never define $\beta(E)$: have i said it right?}
The next-$\lambda$ statistic is $\sF_k$-measurable if we use the LASSO to select variables. \WFcomment{Relate this to spacings test. Is this essentially the ``generalized spacings test?''}
\begin{figure}
  % source code: bivariateSelVSat.R
  \includegraphics[width=\textwidth]{figs/larexample.pdf}
  \label{fig:larexample}
\end{figure}

@@

The task is made much simpler by the fact that the selection event can we written as $S_M=\{y: \Gamma y \geq u\}$.
Hence we don't need to run the entire forward stepwise procedure on $y^*$: instead, we pre-compute $\Gamma$ and $u$ and then check
if $\Gamma y^* \geq u$. Further, since $\Gamma$ has on the  order of $2pk$ rows after $k$ steps,  we can save 
substantially in computation by checking the row-wise inequalities starting from the last row, where the inequality  it is most likely to fail.

@@

\begin{table}[ht]
\centering
\begin{tabular}{rrrrr}
  \hline
  Method, Stopping rule& $p_{\text screen}$ & FWER$_{\text mod}$ & FDR$_{\text mod}$ &FDR$_{\text var}$ \\
% & \$p\_\{$\backslash$text\{screen\}\} & \$$\backslash$text\{FWER\}\_\{$\backslash$text\{model\}\}\$ & \$$\backslash$text\{FDR\}\_\{$\backslash$text\{model\}\}\$ & \$$\backslash$text\{FDR\}\_\{$\backslash$text\{variable\}\}\$ \\ 
  \hline
Nominal, Simple & 0.832 & 0.252 & 0.097 & 0.171 \\ 
  Nominal, Forward & 0.931 & 0.896 & 0.345 & 0.426 \\ 
  Nominal, Strong & 0.003 & 0.000 & 0.000 & 0.035 \\ 
  \hline
  MaxT, Simple & 0.369 & 0.000 & 0.002 & 0.042 \\ 
  MaxT, Forward & 0.668 & 0.038 & 0.026 & 0.080 \\ 
  MaxT, Strong & 0.025 & 0.006 & 0.004 & 0.038 \\ 
  \hline
  MaxT-Identify, Simple & 0.404 & 0.000 & 0.002 & 0.042 \\ 
  MaxT-Identify, Forward & 0.688 & 0.039 & 0.026 & 0.082 \\ 
  MaxT-Identify, Strong & 0.024 & 0.003 & 0.002 & 0.037 \\ 
  \hline
  Saturated, Simple & 0.000 & 0.000 & 0.000 & 0.029 \\ 
  Saturated, Forward & 0.013 & 0.001 & 0.000 & 0.030 \\ 
  Saturated, Strong & 0.000 & 0.000 & 0.000 & 0.033 \\ 
  \hline
  Knockoffs & 0.252 &  &  & 0.103 \\ 
   \hline
\end{tabular}
\caption[tab:stopping]{\em Results of various stopping rules applied to simulated data with 7 strong signals,  as described at the beginning of Section \ref{sec:sparseReg}.
The {\tt Simple} rule stops at the first time that a p-value exceeds $\alpha$, while {\tt Forward} and {\tt Strong} refers ForwardStop and StrongStrop rules of Section \ref{sec:sequential}.}
\label{tab:stopping}
\end{table}

@@

\WFcomment{Can we do this one?}
Here we consider model selection for   principal components analysis. In that case we are given a data matrix $X \in \R^{n\times d}$, with which we form a sample %covariance matrix
\[
S = \frac{1}{n-1} \sum_{i=1}^n(x_i - \bar x)^2
\]
The first $d$ principal component loadings are the first $d$ eigenvectors of $S$, which call $u_1,\ldots, u_d$. These induce a sequence of nested Wishart models:
\[
M_0 \sub M_1 \sub \cdots \sub M_d
\]
in which
\begin{equation}
  M_k:\; (n-1) S \sim W_d\left(\lambda_0 I_d + \sum_{i=1}^k     \lambda_i u_i u_i', \;\;\; n-1\right).
\end{equation}
 By successively conditioning on $u_j, j<k$ one can obtain an exact sequential test
for the complete null hypothesis $H_0: \lambda_j=0 \; \forall j \neq k$.  The computation of the resulting p-values requires numerical integration.
The details are given iin \citet{choi2014selecting}.

@@

We plan to distribute  open source software, in both the R and Python languages, , for computing the max-t and other tests discussed in this paper.

>>>>

\begin{table}[ht]
\centering
\begin{tabular}{rrrrr}
  \hline\\
 Step& Predictor& Naive pvalue& Saturated pvalue& Max-t pvalue\cr
  \hline\\
1 & 3& 0.00 & 0.00 & 0.00 \\ 
  2 & 9 & 0.00 & 0.00 & 0.00 \\ 
  3 & 4 & 0.00 & \color{red}{0.05} & 0.00 \\ 
  4 & 20 & 0.00 & 0.33 & 0.01 \\ 
  5 & 37 & 0.00 & 0.76 & 0.05 \\ 
  6 & 7 & 0.00 & 0.25 & 0.05 \\ 
  7 & 2 & 0.00 & 0.00 & 0.00 \\ 
  8 & 19 & 0.02 & 0.03 & \color{red}{0.37} \\ 
  9 & 11 & 0.11 & 0.55 & 0.95 \\ 
  10 & 52 & 0.21 & 0.91 & 0.92 \\ 
  11 & 57 & 0.05 & 0.37 & 0.21 \\ 
  12 & 24 & 0.22 & 0.15 & 0.00 \\ 
  13 & 5 & 0.18 & 0.07 & 0.10 \\ 
  14 & 49 & 0.16 & 0.97 & 0.75 \\ 
  15 & 30& 0.21 & 0.15 & 0.05 \\ 
  16 & 46 & \color{red}{0.16} & 0.05 & 0.53 \\ 
  17 & 59 & 0.27 & 0.45 & 0.68 \\ 
  18 & 33 & 0.12 & 0.71 & 0.80 \\ 
  19 & 62 & 0.08 & 0.40 & 0.48 \\ 
  20 & 27 & 0.17 & 0.60 & 0.50 \\ 
   \hline\\
\end{tabular}
\caption[tab:diab]{\em Forward stepwise regression for the diabetes data: naive p-values from forward stepwise regression, 
p-values from the ``saturated'' model, and the ``max-t'' p-values that are presented in this paper. The red annotation indicates the stopping step
prescribed by {\tt forwardStop}. For the max-t p-values, this gives exact FDR control at the 10\% level.}
\label{tab:diab}
\end{table}
The table shows the naive p-values from forward stepwise regression, that do not account for the selection, as well as the
p-values from the ``saturated'' model, and the ``max-t'' p-values that are presented in this paper.
The latter two sets of p-values test someone different hypotheses:  the p-values from the saturated model test whether the partial regression of the predictor just added is zero,
what we call the ``incremental null''
In contrast, the max-t pvalues test whether the current model is adequate, that is, all predictors not in the model  have coefficients equal to zero,
the ``complete null''. (The issue of what hypothesis to test is discussed in section \ref{sec:whichnull}.)
Now it turns out that the max-t test is often more powerful that the saturated model test, and importantly, the max-t pvalues are exactly
independent after we reach the point the model sequence when the complete null is true. This independence means that we can apply
sequential FDR exact stopping rules, such as ``forwardStop'' (\citet{gsell2013sequential}). When applied to the max-t pvalues at FDR 0.10, {\tt forwardStop}
suggested stopping at step 8. In constant, the saturated lvalues gave stopping a point of 3, with no guarantee of FDR control
due to the lack of independence of the p-values.

@@

By contrast, the LASSO path does not in general add the variable with the largest $t$-statistic. In that case, we could use instead the {\em selective next-$\lambda$ test}, which uses test statistic
\begin{equation}
  U_k = \inf\{ \lambda:\; \hat\beta^\lambda \notin \beta(E_{k-1})\}.
\end{equation}
Here $\beta(E_{k-1})$ is the support of the ever-active set $E_{k-1}$, so that $U_k$ is the next $\lambda$ at which a new predictor is added by the lasso.
\RTcomment{Will: we never define $\beta(E)$: have i said it right?}
The next-$\lambda$ statistic is $\sF_k$-measurable if we use the LASSO to select variables. \WFcomment{Relate this to spacings test. Is this essentially the ``generalized spacings test?''}

Table \ref{tab:testsummary} summarizes a number of different tests that have been proposed for least angle regression, lasso and forward stepwise regression.
Some focus on the complete null hypothesis, others on the incremental null (as defined in Section \ref{sec:whichnull}).
Some use the saturated model, others the selective model framework, as detailed in earlier in the section.
The saturated model tests are simpler and do not require sampling, while the selective tests are generally more powerful. In the next section we compare the  selective and saturated model settings in more detail.

\begin{table}
\begin{center}
\begin{tabular}{ l | r|r| r}
Procedure& Complete Null & Incremental Null\\
\hline
LAR /lasso& Covariance test/Saturated&Spacing test/Saturated \\
        & Next-$\lambda$ test/Selective& \\
        \hline
Forward Stepwise &maxT, maxT-identify/Selective &Gaussian pivot/Saturated\\
\hline
\end{tabular}
\end{center}
\caption{\em Summary of different tests that have been proposed for least angle regression, lasso and forward stepwise regression.
Tests are described by the null hypothesis that they test (complete or incremental)  and the underlying model (saturated or selective).
The maxT, maxT-identify and next-$\lambda$ tests are proposed in this paper, while the covariance test was proposed in \citet{lockhart2014significance}, and the spacing and Gaussian
pivot tests are given in \citet{spacings}. The latter paper shows that the covariance and spacing test are asymptotically equivalent for the first null step, when the complete null holds,}
\label{tab:testsummary}
\end{table}
Figure  \ref{fig:larexample} shows an example of the LAR/lasso tests with $N=50, p=10$ and two strong signals.
We plot the p-values only for those realizations in which the correct model was selected in the first two steps (about 95\% of the time).
We see that the next-$\lambda$ test is more powerful , especially at step 1. The next-$\lambda$ and spacing test have the correct type I error in steps 3 and 4,
while the covariance test becomes conservative in Step 4, as we move further into the null part of the sequence.
\begin{figure}
  \includegraphics[width=\textwidth]{figs/larexample.pdf}
  \caption[fig:larexample]{\em  Cumulative distribution functions of p-values for LAR/lasso tests from simulations with $N=50,  p=10$ and two strong signals.}
  \label{fig:larexample}
\end{figure}

@@

The task is made much simpler by the fact that the selection event can we written as $S_M=\{y: \Gamma y \geq u\}$.
Hence we don't need to run the entire forward stepwise procedure on $y^*$: instead, we pre-compute $\Gamma$ and $u$ and then check
if $\Gamma y^* \geq u$. Further, since $\Gamma$ has on the  order of $2pk$ rows after $k$ steps,  we can save 
substantially in computation by checking the row-wise inequalities starting from the last row, where the inequality is most likely to fail.

@@

\begin{table}[ht]
\centering
\begin{tabular}{rrrrr}
  \hline
  Method, Stopping rule& $p_{\text screen}$ & FWER$_{\text mod}$ & FDR$_{\text mod}$ &FDR$_{\text var}$ \\
% & \$p\_\{$\backslash$text\{screen\}\} & \$$\backslash$text\{FWER\}\_\{$\backslash$text\{model\}\}\$ & \$$\backslash$text\{FDR\}\_\{$\backslash$text\{model\}\}\$ & \$$\backslash$text\{FDR\}\_\{$\backslash$text\{variable\}\}\$ \\ 
  \hline
Nominal, Simple & 0.832 & 0.252 & 0.097 & 0.171 \\ 
  Nominal, Forward & 0.931 & 0.896 & 0.345 & 0.426 \\ 
  Nominal, Strong & 0.003 & 0.000 & 0.000 & 0.035 \\ 
  \hline
  MaxT, Simple & 0.369 & 0.000 & 0.002 & 0.042 \\ 
  MaxT, Forward & 0.668 & 0.038 & 0.026 & 0.080 \\ 
  MaxT, Strong & 0.025 & 0.006 & 0.004 & 0.038 \\ 
  \hline
  MaxT-Identify, Simple & 0.404 & 0.000 & 0.002 & 0.042 \\ 
  MaxT-Identify, Forward & 0.688 & 0.039 & 0.026 & 0.082 \\ 
  MaxT-Identify, Strong & 0.024 & 0.003 & 0.002 & 0.037 \\ 
  \hline
  Saturated, Simple & 0.000 & 0.000 & 0.000 & 0.029 \\ 
  Saturated, Forward & 0.013 & 0.001 & 0.000 & 0.030 \\ 
  Saturated, Strong & 0.000 & 0.000 & 0.000 & 0.033 \\ 
  \hline
  Knockoffs & 0.252 &  &  & 0.103 \\ 
   \hline
\end{tabular}
\caption[tab:stopping]{\em Results of various stopping rules applied to simulated data with 7 strong signals,  as described at the beginning of Section \ref{sec:sparseReg}.
The {\tt Simple} rule stops at the first time that a p-value exceeds $\alpha$, while {\tt Forward} and {\tt Strong} refers to the ForwardStop and StrongStrop rules of Section \ref{sec:sequential}.}
\label{tab:stopping}
\end{table}

@@

\WFcomment{Can we do this one?}
Here we consider model selection for   principal components analysis. In that case we are given a data matrix $X \in \R^{n\times d}$, with which we form a sample %covariance matrix
\[
S = \frac{1}{n-1} \sum_{i=1}^n(x_i - \bar x)^2
\]
The first $d$ principal component loadings are the first $d$ eigenvectors of $S$, which call $u_1,\ldots, u_d$. These induce a sequence of nested Wishart models:
\[
M_0 \sub M_1 \sub \cdots \sub M_d
\]
in which
\begin{equation}
  M_k:\; (n-1) S \sim W_d\left(\lambda_0 I_d + \sum_{i=1}^k     \lambda_i u_i u_i', \;\;\; n-1\right).
\end{equation}
 By successively conditioning on $u_j, j<k$ one can obtain an exact sequential test
for the complete null hypothesis $H_0: \lambda_j=0 \; \forall j \neq k$.  The computation of the resulting p-values requires numerical integration.
The details are given in \citet{choi2014selecting}, with their ``integrated CSV'' test  corresponding to the sequential selective test of this paper.

@@

We plan to distribute  open source software, in both the R and Python languages,  for computing the max-t and other tests discussed in this paper.
